Why do people punish? Evidence for a range of strategic concerns

### 2024-07-24 manuscript.tex
submit manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{punishment; cooperation; economic games\newline\indent Word count: 5657 words}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{makecell}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\nolinenumbers
\note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Why do people punish? Evidence for a range of strategic concerns},
  pdfauthor={Scott Claessens*1, Quentin D. Atkinson1, \& Nichola Raihani1,2},
  pdflang={en-EN},
  pdfkeywords={punishment; cooperation; economic games},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\nolinenumbers

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One
mechanism thought to sustain this level of cooperation is costly punishment,
whereby individuals harm others at a personal cost\textsuperscript{1},
ostensibly encouraging cooperative behaviour from the target (or bystanders\textsuperscript{2--4}) in the future. Punishment offers a
route to maintaining or increasing cooperation by changing the payoff structure
of social interactions such that it no longer pays to cheat or exploit social
partners\textsuperscript{1,5}. Yet, despite its theoretical
importance, the question of why people choose to punish others is still hotly
contested\textsuperscript{6}. In this study, we use a battery of economic games to
disentangle the different motives underpinning punishment and explore how these
motives vary across individuals.

@@

Beyond behaviour shaping concerns, there are a host of other reasons why people
may want to punish in economic games. Punishers might be motivated by a desire
for retribution rather than deterrence, punishing in proportion to the amount
of harm that was personally caused\textsuperscript{28}. Punishment might be driven
by concerns about relative payoffs, such as disadvantageous inequity aversion
(i.e., avoiding having less than others\textsuperscript{6,29}) and/or
general egalitarian preferences (i.e., wanting all participants to receive the
same payoffs\textsuperscript{30}). Such concerns about relative payoffs may be activated
when participants earn less than cheats in economic games or when there are
income disparities in these settings. People might also use punishment for
competitive purposes, seeking advantageous inequity for themselves (i.e., having
more than others) and/or improving their relative position\textsuperscript{6}.

Common economic game designs have been unable to tease apart these different
motives for punishment because participants who interact with cheats in these
games experience both losses \emph{and} lower relative payoffs. The typical 1:3
fee-fine ratio of punishment in economic games compounds this issue. With
this setup, people can simultaneously use punishment to reciprocate losses, to
deter others from cheating, and to reduce or reverse disparities in payoffs
between themselves and targets. To add to this complexity, it is evident that
people use punishment in seemingly disparate ways: punishing when no behaviour
change is possible, such as in one-shot games\textsuperscript{13,29,31,32}, on the very last round of repeated games\textsuperscript{33}, or in games where the target never learns about the
punishment\textsuperscript{34}; punishing those who did not cheat or who
over-contributed to collective action (antisocial punishment\textsuperscript{17,35}); punishing in scenarios where they were not
personally harmed (third-party punishment\textsuperscript{36}); and punishing in
scenarios where disparities in payoffs did not arise from participants'
actions\textsuperscript{30,37,38}.

@@

Demographic variables tended to be unrelated to strategy usage: age and gender
did not predict adherence to a particular punishment strategy (Supplementary
Figures \ref{fig:plotAllDems2} and \ref{fig:plotAllDems1}). In the United
States, the never punish strategy was slightly more common among participants
lower in socio-economic status (median posterior slope =
-0.20, 95\% CI {[}-0.38
-0.02{]}) but this effect was small.

@@

We defined antisocial punishment as occurring when a participant exclusively
punished those who did not steal. There were no participants in our sample who
followed this strategy, although some players did punish non-stealing
co-players. Harming non-stealing individuals was also consistent with the
competitive strategies, which did appear in our sample albeit at low frequencies.
The fact that no participants in our sample punished non-stealing across all
games suggests that the antisocial punishment that has been observed in other
studies\textsuperscript{17,51} is not aimed at harming cooperators
specifically, as has been previously suggested\textsuperscript{17}. Instead,
antisocial punishment is more likely to be motivated by improving one's
relative position, which is in line with work showing that antisocial punishment
disappears when relative payoffs cannot be changed (e.g., when punishment is
only available with a 1:1 fee-fine ratio\textsuperscript{35}).

@@

It is striking that the most common strategy in our dataset was to never punish.
This is partly because punishment in these games imposes an economic cost for no
tangible benefit. If the fee-fine ratio had been lower, such that it was cheaper
to punish, we may have seen more punishment from participants. Indeed,
72\%
of participants following the never punish strategy positively stated that they
didn't want to pay to reduce anyone's bonus but would have done so if it were
free. But the frequency of the never punish strategy perhaps also reflects a
more general aversion to peer punishment, an aversion that has been highlighted
in both WEIRD (Western, educated, industrialised, rich, and democratic) samples\textsuperscript{53,54} and in small-scale societies\textsuperscript{55}.
One reason that people may be averse to peer punishment is that, because it is
a fundamentally harmful act, punishment can reflect badly on the punisher and
people might therefore refrain from punishing others to avoid reputational
damage\textsuperscript{4}. People frequently avoid taking actions that could harm
their reputation, even when they don't know if reputation is at stake (as in the
one-shot anonymous settings used here). Another reason that people may be averse
to peer punishment is that it can trigger retaliation\textsuperscript{6}. This may be
especially likely in situations that lack clear institutional norms to
legitimise punishment, such as our economic games. As with reputation damage,
people might abstain from peer punishment to avoid retaliation, regardless of
whether retaliation is actually possible. By contrast, institutionalised
punishment in small-scale societies often functions to compensate victims while
limiting the potential for feuds and cycles of retaliation\textsuperscript{56,57}. Future research should uncover whether people are
more willing to punish in these conventionalised contexts (e.g., see\textsuperscript{58}).

We finally highlight some key limitations that can guide future research. One
limitation with our design is that some strategies required more punishment than
others to be met. For example, the competitive strategy required punishment in
ten of twelve decisions, compared to the avoid disadvantageous inequity strategy
which required only one instance of punishment (Table \ref{tab:tableStrategies}).
Strategies thus differed in how ``expensive'' they were to implement, perhaps
explaining why some strategies were more common than others. This issue is
largely unavoidable in our design since strategies, by their very nature, differ
in how punitive they are. To partially mitigate this issue, we employed the
strategy method to incentivise participants, such that payoffs were calculated
from a randomly chosen game instead of summed across all games.

Another limitation is that our results may be contingent on the particular suite
of anonymous stealing games that we used. With our anonymous design, we were
unable to study other potential reputational strategies underlying punishment,
such as signalling trustworthiness\textsuperscript{47}. Moreover, stealing may be
evaluated differently to other forms of cheating, such as not contributing to
public goods, and other negative behaviours, such as lying or breaking taboos.
Finally, we were unable to include all permutations of situational features in
the games (e.g., second-party vs.~third-party, equal vs.~unequal) making it
difficult to interpret some patterns of behaviour. For example, many
participants punished only in the third-party game (Game F), but it is not clear
whether these participants were driven by the third-party nature of the game or
simply by the fact that stealing in that game generated inequality. Future work
should determine whether different reputational contexts, target behaviours, and
combinations of situational features elicit different punishment strategies.

@@

We pre-registered that we would use a Bayesian latent state model to infer
unobserved punishment strategies from the observed data (for a similar version
of this model, see ref\textsuperscript{65}). In this model, participants \(i\) in
countries \(c\) make binary punishment decisions across twelve decisions \(j\). We
assume that the probability of the observed data \(y_{i,j}\) is the weighted
average of the probability of the observed data conditional on each of the ten
punishment strategies \(s\). From this logic, the model estimates the probability
of each strategy \(p_{s}\). The full model is as follows:

@@

\begingroup

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-CluttonBrock1995}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Clutton-Brock, T. H. \& Parker, G. A. \href{https://doi.org/10.1038/373209a0}{Punishment in animal societies}. \emph{Nature} \textbf{373}, 209--216 (1995).}

\leavevmode\vadjust pre{\hypertarget{ref-DosSantos2011}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{dos Santos, M., Rankin, D. J. \& Wedekind, C. \href{https://doi.org/10.1098/rspb.2010.1275}{The evolution of punishment through reputation}. \emph{Proceedings of the Royal Society B: Biological Sciences} \textbf{278}, 371--377 (2011).}

\leavevmode\vadjust pre{\hypertarget{ref-DosSantos2013}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{dos Santos, M., Rankin, D. J. \& Wedekind, C. \href{https://www.jstor.org/stable/24032701}{Human cooperation based on punishment reputation}. \emph{Evolution} \textbf{67}, 2446--2450 (2013).}

\leavevmode\vadjust pre{\hypertarget{ref-Raihani2015}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Raihani, N. J. \& Bshary, R. \href{https://doi.org/10.1016/j.tree.2014.12.003}{The reputation of punishers}. \emph{Trends in Ecology \& Evolution} \textbf{30}, 98--103 (2015).}

\leavevmode\vadjust pre{\hypertarget{ref-Ostrom1990}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Ostrom, E. \emph{Governing the commons: The evolution of institutions for collective action}. (Cambridge University Press, 1990).}

\leavevmode\vadjust pre{\hypertarget{ref-Raihani2019}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Raihani, N. J. \& Bshary, R. \href{https://doi.org/10.1017/ehs.2019.12}{Punishment: One tool, many uses}. \emph{Evolutionary Human Sciences} \textbf{1}, e12 (2019).}

\leavevmode\vadjust pre{\hypertarget{ref-Balliet2011}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Balliet, D., Mulder, L. B. \& Van Lange, P. A. M. \href{https://doi.org/10.1037/a0023489}{Reward, punishment, and cooperation: A meta-analysis}. \emph{Psychological Bulletin} \textbf{137}, 594--615 (2011).}

\leavevmode\vadjust pre{\hypertarget{ref-Balliet2013}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Balliet, D. \& Lange, P. A. M. V. \href{https://doi.org/10.1177/1745691613488533}{Trust, punishment, and cooperation across 18 societies: A meta-analysis}. \emph{Perspectives on Psychological Science} \textbf{8}, 363--379 (2013).}

\leavevmode\vadjust pre{\hypertarget{ref-Chaudhuri2011}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Chaudhuri, A. \href{https://doi.org/10.1007/s10683-010-9257-1}{Sustaining cooperation in laboratory public goods experiments: A selective survey of the literature}. \emph{Experimental Economics} \textbf{14}, 47--83 (2011).}

\leavevmode\vadjust pre{\hypertarget{ref-Dreber2008}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Dreber, A., Rand, D. G., Fudenberg, D. \& Nowak, M. A. \href{https://doi.org/10.1038/nature06723}{Winners don't punish}. \emph{Nature} \textbf{452}, 348--351 (2008).}

\leavevmode\vadjust pre{\hypertarget{ref-Fehr2000}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Fehr, E. \& Gächter, S. \href{https://www.jstor.org/stable/117319}{Cooperation and punishment in public goods experiments}. \emph{American Economic Review} \textbf{90}, 980--994 (2000).}

\leavevmode\vadjust pre{\hypertarget{ref-Fehr2002}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Fehr, E. \& Gächter, S. \href{https://doi.org/10.1038/415137a}{Altruistic punishment in humans}. \emph{Nature} \textbf{415}, 137--140 (2002).}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man, donotrepeattitle,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Why do people punish? Evidence for a range of strategic concerns},
  pdfauthor={Scott Claessens*1, Quentin D. Atkinson1, \& Nichola Raihani1,2},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{makecell}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\nolinenumbers
\note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This manuscript has not yet been peer-reviewed. \par This manuscript was published as a pre-print on PsyArXiv under a CC BY 4.0 \text{license:} \url{https://doi.org/10.31234/osf.io/ys6rm} \par \text{Classification:} Social Sciences; Psychological and Cognitive Sciences \par \text{Keywords:} punishment; cooperation; economic games}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

@@

\linenumbers

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One
mechanism thought to sustain this level of cooperation is costly punishment,
whereby individuals harm others at a personal cost\textsuperscript{1},
ostensibly encouraging cooperative behaviour from the target, or bystanders\textsuperscript{2--4}, in the future. Punishment offers a
route to maintaining or increasing cooperation by changing the payoff structure
of social interactions such that it no longer pays to cheat or exploit social
partners\textsuperscript{1,5}. Yet, despite its theoretical
importance, the question of why people choose to punish others is still hotly
contested\textsuperscript{6}. In this study, we use a battery of economic games to
disentangle the different motives underpinning punishment and explore how these
motives vary across individuals.

@@

Beyond behaviour shaping concerns, there are a host of other reasons why people
may want to punish in economic games. Punishers might be motivated by a desire
for retribution rather than deterrence, punishing in proportion to the amount
of harm that was personally caused\textsuperscript{28}. Punishment might be driven
by concerns about relative payoffs, such as disadvantageous inequity aversion
(i.e., avoiding having less than others)\textsuperscript{6,29} and/or
general egalitarian preferences (i.e., wanting all participants to receive the
same payoffs)\textsuperscript{30}. Such concerns about relative payoffs may be activated
when participants earn less than cheats in economic games or when there are
income disparities in these settings. People might also use punishment for
competitive purposes, seeking advantageous inequity for themselves (i.e., having
more than others) and/or improving their relative position\textsuperscript{6}.

Common economic game designs have been unable to tease apart these different
motives for punishment because participants who interact with cheats in these
games experience both losses \emph{and} lower relative payoffs. The typical 1:3
fee-fine ratio of punishment in economic games compounds this issue. With
this setup, people can simultaneously use punishment to reciprocate losses, to
deter others from cheating, and to reduce or reverse disparities in payoffs
between themselves and targets. To add to this complexity, it is evident that
people use punishment in seemingly disparate ways: punishing when no behaviour
change is possible, such as in one-shot games\textsuperscript{13,29,31,32}, on the very last round of repeated games\textsuperscript{33}, or in games where the target never learns about the
punishment\textsuperscript{34}; punishing those who did not cheat or who
over-contributed to collective action (antisocial punishment)\textsuperscript{17,35}; punishing in scenarios where they were not
personally harmed (third-party punishment)\textsuperscript{36}; and punishing in
scenarios where disparities in payoffs did not arise from participants'
actions\textsuperscript{30,37,38}.

@@

Demographic variables tended to be unrelated to strategy usage: age and gender
did not predict adherence to a particular punishment strategy (Supplementary
Figures \ref{fig:plotAllDems2} and \ref{fig:plotAllDems1}). In the United
States, the never punish strategy was slightly more common among participants
lower in socio-economic status (median posterior slope =
-0.20, 95\% CI {[}-0.38
-0.03{]}) but this effect was small.

@@

We defined antisocial punishment as occurring when a participant exclusively
punished those who did not steal. There were no participants in our sample who
followed this strategy, although some players did punish non-stealing
co-players. Harming non-stealing individuals was also consistent with the
competitive strategies, which did appear in our sample albeit at low frequencies.
The fact that no participants in our sample punished non-stealing across all
games suggests that the antisocial punishment that has been observed in other
studies\textsuperscript{17,51} is not aimed at harming cooperators
specifically, as has been previously suggested\textsuperscript{17}. Instead,
antisocial punishment is more likely to be motivated by improving one's
relative position, which is in line with work showing that antisocial punishment
disappears when relative payoffs cannot be changed (e.g., when punishment is
only available with a 1:1 fee-fine ratio)\textsuperscript{35}.

@@

It is striking that the most common strategy in our dataset was to never punish.
This is partly because punishment in these games imposes an economic cost for no
tangible benefit. If the fee-fine ratio had been lower, such that it was cheaper
to punish, we may have seen more punishment from participants. Indeed,
72\%
of participants following the never punish strategy positively stated that they
didn't want to pay to reduce anyone's bonus but would have done so if it were
free. But the frequency of the never punish strategy perhaps also reflects a
more general aversion to peer punishment, an aversion that has been highlighted
in both WEIRD (Western, educated, industrialised, rich, and democratic) samples\textsuperscript{53,54} and in small-scale societies\textsuperscript{55}.
One reason that people may be averse to peer punishment is that, because it is
a fundamentally harmful act, punishment can reflect badly on the punisher and
people might therefore refrain from punishing others to avoid reputational
damage\textsuperscript{4}. People frequently avoid taking actions that could harm
their reputation, even when they don't know if reputation is at stake (as in the
one-shot anonymous settings used here). Another reason that people may be averse
to peer punishment is that it can trigger retaliation\textsuperscript{6}. This may be
especially likely in situations that lack clear institutional norms to
legitimise punishment, such as our economic games. As with reputation damage,
people might abstain from peer punishment to avoid retaliation, regardless of
whether retaliation is actually possible. By contrast, institutionalised
punishment in small-scale societies often functions to compensate victims while
limiting the potential for feuds and cycles of retaliation\textsuperscript{56,57}. Future research should uncover whether people are
more willing to punish in these conventionalised contexts (e.g., see ref\textsuperscript{58}).

One potential limitation with our design is that some strategies required more
punishment than others, meaning that some strategies were more ``expensive'' to
implement. For example, the competitive strategy required punishment in ten of
twelve decisions, compared to the avoid disadvantageous inequity strategy which
required only one instance of punishment (Table \ref{tab:tableStrategies}).
We employed the strategy method to deal with this, calculating participant
payoffs from a randomly chosen game instead of accumulating the costs across
all games. This mitigates the concern that the total cost is the most salient
feature determining participants' decisions. But even accounting for this,
one could still argue that the difference in overall costs explains why, for
example, the competitive strategy is less common in our dataset than the avoid
disadvantageous inequity strategy. We do not think this feature of our design
is a concern, however, for a number of reasons. First, we are interested in
measuring strategies underlying \emph{costly} punishment. Some of these strategies,
by their very nature, will be more costly to implement than others. This is
reflected in our design. It would not make theoretical sense to use an
alternative design where more or less punitive strategies are manipulated to
cost participants the same amount. Second, when we asked participants whether
they would have punished if it were free to do so, we found low agreement
with this statement (Supplementary Figure \ref{fig:plotSliders1}),
suggesting that participants were not particularly sensitive to the costs of
punishment. In line with this, when we plot the frequencies of different
strategies against their expected costs, we find that cost is not a perfect
predictor of strategy frequency: many ``cheap'' strategies are rare in our
data and some ``expensive'' strategies are quite common (e.g., always punish;
Supplementary Figure \ref{fig:plotSpline}). Third, this argument implies
that the high frequency of the never punish strategy is merely an artefact
of our design, since it is the only strategy that does not cost anything to
implement. But as we have discussed, many other studies have found similar
aversions to costly punishment in the lab\textsuperscript{29,31,33,44,48} and in the real world\textsuperscript{53--55} suggesting that this result is
not an artefact of our design.

@@

We pre-registered that we would use a Bayesian latent state model to infer
unobserved punishment strategies from the observed data {[}for a similar version
of this model, see ref\textsuperscript{65}{]}. In this model, participants \(i\) in
countries \(c\) make binary punishment decisions across twelve decisions \(j\). We
assume that the probability of the observed data \(y_{i,j}\) is the weighted
average of the probability of the observed data conditional on each of the ten
punishment strategies \(s\). From this logic, the model estimates the probability
of each strategy \(p_{s}\). The full model is as follows:

@@

\begingroup
\setlength\parindent{0pt}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-CluttonBrock1995}{}%
1. Clutton-Brock, T. H. \& Parker, G. A. Punishment in animal societies. \emph{Nature} \textbf{373}, 209--216 (1995).

\leavevmode\hypertarget{ref-DosSantos2011}{}%
2. dos Santos, M., Rankin, D. J. \& Wedekind, C. The evolution of punishment through reputation. \emph{Proceedings of the Royal Society B: Biological Sciences} \textbf{278}, 371--377 (2011).

\leavevmode\hypertarget{ref-DosSantos2013}{}%
3. dos Santos, M., Rankin, D. J. \& Wedekind, C. Human cooperation based on punishment reputation. \emph{Evolution} \textbf{67}, 2446--2450 (2013).

\leavevmode\hypertarget{ref-Raihani2015}{}%
4. Raihani, N. J. \& Bshary, R. The reputation of punishers. \emph{Trends in Ecology \& Evolution} \textbf{30}, 98--103 (2015).

\leavevmode\hypertarget{ref-Ostrom1990}{}%
5. Ostrom, E. \emph{Governing the commons: The evolution of institutions for collective action}. (Cambridge University Press, 1990).

\leavevmode\hypertarget{ref-Raihani2019}{}%
6. Raihani, N. J. \& Bshary, R. Punishment: One tool, many uses. \emph{Evolutionary Human Sciences} \textbf{1}, e12 (2019).

\leavevmode\hypertarget{ref-Balliet2011}{}%
7. Balliet, D., Mulder, L. B. \& Van Lange, P. A. M. Reward, punishment, and cooperation: A meta-analysis. \emph{Psychological Bulletin} \textbf{137}, 594--615 (2011).

\leavevmode\hypertarget{ref-Balliet2013}{}%
8. Balliet, D. \& Lange, P. A. M. V. Trust, punishment, and cooperation across 18 societies: A meta-analysis. \emph{Perspectives on Psychological Science} \textbf{8}, 363--379 (2013).

\leavevmode\hypertarget{ref-Chaudhuri2011}{}%
9. Chaudhuri, A. Sustaining cooperation in laboratory public goods experiments: A selective survey of the literature. \emph{Experimental Economics} \textbf{14}, 47--83 (2011).

\leavevmode\hypertarget{ref-Dreber2008}{}%
10. Dreber, A., Rand, D. G., Fudenberg, D. \& Nowak, M. A. Winners don't punish. \emph{Nature} \textbf{452}, 348--351 (2008).

\leavevmode\hypertarget{ref-Fehr2000}{}%
11. Fehr, E. \& Gächter, S. Cooperation and punishment in public goods experiments. \emph{American Economic Review} \textbf{90}, 980--994 (2000).

\leavevmode\hypertarget{ref-Fehr2002}{}%
12. Fehr, E. \& Gächter, S. Altruistic punishment in humans. \emph{Nature} \textbf{415}, 137--140 (2002).

\leavevmode\hypertarget{ref-Henrich2006}{}%
13. Henrich, J. \emph{et al.} Costly punishment across human societies. \emph{Science} \textbf{312}, 1767--1770 (2006).

\leavevmode\hypertarget{ref-Nikiforakis2008a}{}%
14. Nikiforakis, N. \& Normann, H.-T. A comparative statics analysis of punishment in public-good experiments. \emph{Experimental Economics} \textbf{11}, 358--369 (2008).

\leavevmode\hypertarget{ref-Raihani2012a}{}%
15. Raihani, N. J., Thornton, A. \& Bshary, R. Punishment and cooperation in nature. \emph{Trends in Ecology \& Evolution} \textbf{27}, 288--295 (2012).

\leavevmode\hypertarget{ref-DeQuervain2004}{}%
16. de Quervain, D. J.-F. \emph{et al.} The neural basis of altruistic punishment. \emph{Science} \textbf{305}, 1254--1258 (2004).

\leavevmode\hypertarget{ref-Herrmann2008}{}%
17. Herrmann, B., Thöni, C. \& Gächter, S. Antisocial punishment across societies. \emph{Science} \textbf{319}, 1362--1367 (2008).

\leavevmode\hypertarget{ref-Camerer2003}{}%
18. Camerer, C. F. \emph{Behavioral game theory: Experiments in strategic interaction}. (Russell Sage Foundation, 2003).

\leavevmode\hypertarget{ref-Bowles2013}{}%
19. Bowles, S. \& Gintis, H. \emph{A cooperative species: Human reciprocity and its evolution}. (Princeton University Press, 2013).

\leavevmode\hypertarget{ref-Boyd2003}{}%
20. Boyd, R., Gintis, H., Bowles, S. \& Richerson, P. J. The evolution of altruistic punishment. \emph{Proceedings of the National Academy of Sciences} \textbf{100}, 3531--3535 (2003).

\leavevmode\hypertarget{ref-Chudek2011}{}%
21. Chudek, M. \& Henrich, J. Culture-gene coevolution, norm-psychology and the emergence of human prosociality. \emph{Trends in Cognitive Sciences} \textbf{15}, 218--226 (2011).
